Climate Change Adaptation: Uncertainty Associated with Sequential Finite Period Decisions under Nonstationarity

### 2018-02-14 writeup/Outline/main.tex
write up the methods

\subsection{Nonstationary Flood Frequency Analysis}

@@

We generate synthetic flood sequences from a simple conceptual model where the expected flood varies as a function of a single climate index plus time following the log-normal distribution:
\begin{equation} \label{eq:lognormal}
  \log Q(t) \big| \mu(t),\sigma(t) \sim \normal \qty(\mu(t), \sigma(t)).
\end{equation}

ENSO is the dominant mode of tropical climate variability, impacts flood risk around the world, and varies in a quasi-oscillatory manner with variability on the order of \SIrange{4}{6}{year} modulated by low-frequency oscillation \citep[see][for a comprehensive review]{Sarachik2010}.
A \SI{100000}{year} integration of the Cane-Zebiak model \citep{Zebiak1987} was used to produce a monthly NINO3.4 index with stationary forcing as described in \citep{Ramesh2017}.
To create an annual time series, we average the October-December months to create a time series \( x(t) \) where \( t \) is taken in annual steps \(1, \ldots, \num{100000} \).
The parameters of the distribution vary in time according to the sampled NINO3 index denoted \( x(t) \) as well as time itself, representing secular trends.
First a starting point in the \SI{100000}{year} ENSO was randomly chosen, then \(\mu\) is determined by
\begin{equation}
  \mu(t) = \mu_0 + \beta x(t) + \gamma \qty(t - t_0)
\end{equation}
and a constant coefficient of variation
\begin{equation}
  \sigma(t) = \begin{cases} \alpha \mu(t) & \qqtext{if} \alpha \mu(t) \geq \sigma_\text{min} \\ \sigma_\text{min} & \qqtext{else} \end{cases}
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.

Though this is a highly simplified model neglecting many important mechanisms, it enables us to generate many synthetic sequences of \(\qty[M + N]\) years which exhibit chaotic but organized quasi-periodic behavior (regime behavior or low-frequency variability) plus a trend term, thereby capturing many essential features of hydrological variability.

@@

\subsection{Fitting Streamflow Sequences\label{sec:estimation}}

@@

\subsection{Hidden Markov Model\label{sec:HMM}}

@@

\begin{table}[b]
  \begin{center}
    \begin{tabular}{L{1.0in} L{3.0in} L{1.5in}}
      \toprule
        Model Name & Description & Relevant Section \\
      \midrule
        stationary & \(Q \sim \GEV \qty(\mu, \sigma, \xi)\) & \cref{sec:estimation} \\
        location trend & \(Q \sim \GEV \qty(\mu_0 + \beta_\mu \qty(t-t_0), \sigma, \xi)\) & \cref{sec:estimation} \\
        HMM & log-normal Hidden Markov Model & \cref{sec:HMM} \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{Summary of models used for fitting\label{tab:model-fitting}}
\end{table}

\subsection{Flood Estimation and Uncertainty}

@@

\section{Are Generated Sequences Realistic?\label{sec:sequence-realistic}}

@@

\section{Stationary, Static Risk}

Consider static risk and a stationary process, and partiuclarly implications of \(\qty{M, N}\).
Much of this is already laid out in \citet{Lall1987}.

\section{Stationary, Static Risk}

Consider static risk and a non-stationary process.
Estimate using drift or nonstationary models using full \(N\) of the most recent \(N\) (\ie{} what Vogel is proposing).
Illustrate via simulation for different signal to noise ratio of nonstationary terms and \(\qty{M, N}\).

\section{Dynamic Risk}

Consider dynamic risk, \ie{} updating and sequential decisions, a nonstationary generative process, and \(M\) and \(N\) considerations.

@@

Hopefully something to discuss?

For the type of extreme flood events we are interested in, the nominal return period at some location flooded during the event will typically be larger than 10 years. 
While this is a relatively rare event at a given location, across a geography as large as the United States, one can expect to see such events rather frequently, emphasizing the need to understand and predict the risk of such events over a season or few years to facilitate preparation and mitigation strategies.\footnote{Send this to discussion}

@@

\section{Notation}

\begin{description}
  \item[\(Q(t)\)] Annual-maximum flood series
  \item[\(Q_T\)] Flood threshold
\end{description}

>>>>

\subsection{Nonstationary Flood Frequency Analysis\label{sec:ffa}}

@@

We generate synthetic flood sequences from a simple conceptual model where the expected flood varies as a function of a single climate index plus time following the log-normal distribution:
\begin{equation} \label{eq:lognormal}
  \log Q(t) \big| \mu(t),\sigma(t) \sim \normal \qty(\mu(t), \sigma(t)).
\end{equation}
The parameter \( \mu(t) \) is generated using two different approaches, summarized below, while for both we assume a constant coefficient of variation:
\begin{equation}
  \sigma(t) = \begin{cases} \alpha \mu(t) & \qqtext{if} \alpha \mu(t) \geq \sigma_\text{min} \\ \sigma_\text{min} & \qqtext{else} \end{cases}
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.
Though both approaches used to generate sequences \( \mu(t) \) use ighly simplified models neglecting many important mechanisms, they enable us to generate many synthetic sequences of \(\qty[M + N]\) years which exhibit chaotic but organized quasi-periodic behavior (regime behavior or low-frequency variability) plus a trend term, thereby capturing many essential features of hydrological variability.

@@

\subsubsection{Synthetic ENSO Model\label{methods-enso}}

ENSO is the dominant mode of tropical climate variability, impacts flood risk around the world, and varies in a quasi-oscillatory manner with variability on the order of \SIrange{4}{6}{year} modulated by low-frequency oscillation \citep[see][for a comprehensive review]{Sarachik2010}.
A \SI{100000}{year} integration of the Cane-Zebiak model \citep{Zebiak1987} was used to produce a monthly NINO3.4 index with stationary forcing as described in \citep{Ramesh2017}.
To create an annual time series, we average the October-December months to create a time series \( x(t) \) where \( t \) is taken in annual steps \(1, \ldots, \num{100000} \).
The parameters of the distribution vary in time according to the sampled NINO3 index denoted \( x(t) \) as well as time itself, representing secular trends.
First a starting point in the \SI{100000}{year} ENSO was randomly chosen, then \(\mu\) is determined by
\begin{equation}
  \mu(t) = \mu_0 + \beta x(t) + \gamma \qty(t - t_0)
\end{equation}
where \(\sigma_\text{min}\) is a small constant that ensures non-trivial noise in the system.

\subsubsection{Markov Chain}

Alternatively, we consider a 2-state Markov chain with a constant transition probability which is the same for both states, such that the transition matrix is
\begin{equation*}
  P = \mqty[\pi & 1-\pi \\ 1-\pi & \pi]
\end{equation*}
This is first used to generate a sequence of states \(S(t)\).
The value \( \mu(t) \) depends only on the \(S(t)\) and the time itself:
\begin{equation*}
  \mu(t) = \begin{cases}
    \mu_{0,1} + \gamma_1 \qty(t - t_0) & \qqtext{if} S(t) = 1 \\
    \mu_{0,2} + \gamma_2 \qty(t - t_0) & \qqtext{if} S(t) = 2
  \end{cases}
\end{equation*}
For simplicity, we assume that the coefficient of variation \( \alpha \) is the same for both states.

\subsection{Estimating Future Streamflow Sequences\label{sec:estimation}}

We do not, in general, know the true distribution of annual-maximum floods, and as discussed in \cref{sec:ffa} a plethora of approaches have been suggested in the literature.
This paper does not propose new models or suggest a general ``best'' model; instead it aims to describe the total uncertainty in estimates using easily methods which have been well studied in the literature.

In general, each historical \(N\)-year sequence of streamflow is fit to a statistical model, from which 1000\footnote{Check this parameter} sequences of future streamflow are generated using Monte Carlo simulation.
These estimates are then used to calculate \( \hat{p_T} \), and the estimates are evaluated as described in \cref{sec:methods-uncertainty}.
Three models are assessed for estimating future streamflow; they are summarized in \cref{tab:model-fitting} and described in greater detail below.

\begin{table}[bht]
  \begin{center}
    \begin{tabular}{L{1.0in} L{3.0in} L{1.5in}}
      \toprule
        Model Name & Description & Relevant Section \\
      \midrule
        stationary & \(Q \sim \GEV \qty(\mu, \sigma, \xi)\) & \cref{sec:method-stationary} \\
        location trend & \(Q \sim \GEV \qty(\mu_0 + \beta_\mu \qty(t-t_0), \sigma, \xi)\) & \cref{sec:method-trend} \\
        HMM & log-normal Hidden Markov Model & \cref{sec:method-HMM} \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{Summary of models used for fitting\label{tab:model-fitting}}
\end{table}

\subsubsection{Stationary Model\label{sec:method-stationary}}

@@

\subsubsection{Trend Model\label{sec:method-trend}}

As \cref{sec:method-stationary} except that
\begin{equation*}
  \mu = \mu_0 + \beta_\mu \qty(t - t_0)
\end{equation*}
and so the parameters \( \mu_0, \beta_\mu \) must be estimated rather than just \(\mu\).

\subsubsection{Hidden Markov Model\label{sec:method-HMM}}

@@

\subsection{Flood Estimation and Uncertainty\label{sec:methods-uncertainty}}

@@

\section{Results}

\subsection{Are Generated Sequences Realistic?\label{sec:sequence-realistic}}

@@

\subsection{Stationary Case}

\subsection{Trend Case}

\section{Case Study}

We still need to find a good case study -- it can be a highly simplified one.
We can come up with some numbers and a single streamflow sequence that has been observed.

@@

For the type of extreme flood events we are interested in, the nominal return period at some location flooded during the event will typically be larger than 10 years.
While this is a relatively rare event at a given location, across a geography as large as the United States, one can expect to see such events rather frequently, emphasizing the need to understand and predict the risk of such events over a season or few years to facilitate preparation and mitigation strategies.

@@

\section*{Notation}

\begin{description}
  \item[\( Q(t) \)] Annual-maximum flood series
  \item[\( Q_T \)] Flood threshold
  \item[\( M \)] the project planning period
  \item[\( N \)] the length of the observational record, in years
  \item[\( p_T \)] the true probability that \( Q(t) \geq Q_T \) in the future \(M\)-year period
  \item[\( \hat{p}_T \)] the estimated probability that \( Q(t) \geq Q_T \) in the future \(M\)-year period
\end{description}
